---
title: "Fast donor-weighted DE for multi-donor scRNA-seq with scFastDE"
author:
  - name: "Your Name"
    affiliation: "Your Institution"
    email: "your@email.com"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Fast donor-weighted DE for multi-donor scRNA-seq with scFastDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse  = TRUE,
    comment   = "#>",
    fig.width = 7,
    fig.height = 5,
    message   = FALSE,
    warning   = FALSE
)
```

# Introduction

Multi-donor single-cell RNA-seq studies are now the norm for disease studies,
clinical trials, and population-level atlases. The standard approach to
differential expression (DE) in these datasets is pseudo-bulk analysis:
aggregate per-cell counts into per-donor profiles, then apply bulk DE tools
like `DESeq2` or `edgeR`.

This works well but has two underappreciated problems:

**Problem 1: Speed.** Bulk DE tools fit one model per gene in a serial R
loop. On a dataset with 30,000 genes and 50 donors, this loop runs 30,000
times. At scale — 100k+ cells, multiple cell types — this becomes the
analysis bottleneck.

**Problem 2: Donor weighting.** Standard pseudo-bulk tools treat a donor
with 5 cells identically to a donor with 500 cells. The 5-cell pseudo-bulk
is far noisier, yet gets equal weight in the DE model.

`scFastDE` addresses both:

1. **Vectorised testing** — the entire pseudo-bulk matrix is passed to
   `limma::lmFit` in one call, which uses LAPACK routines to fit all gene
   models simultaneously.
2. **Cell-count weights** — each donor is weighted by `sqrt(n_cells)`,
   giving principled influence to well-represented donors.
3. **Sparse donor guard** — donors with too few cells are removed before
   aggregation, preventing noisy pseudo-bulks from distorting results.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("scFastDE")
```

# Quick start

```{r quickstart}
library(scFastDE)
library(SingleCellExperiment)

# Simulate a multi-donor, two-condition dataset
set.seed(2024)
n_genes <- 500
n_cells <- 120

counts <- matrix(rpois(n_genes * n_cells, 8L),
                 nrow = n_genes, ncol = n_cells)
rownames(counts) <- paste0("Gene", seq_len(n_genes))
colnames(counts) <- paste0("Cell", seq_len(n_cells))

# Inject DE signal: first 20 genes upregulated in treatment
counts[1:20, 61:120] <- counts[1:20, 61:120] * 4L

sce <- SingleCellExperiment(assays = list(counts = counts))
sce$donor     <- rep(paste0("Donor", 1:12), each = 10)
sce$cell_type <- "CD4_Tcell"
sce$condition <- rep(c("ctrl", "treat"), each = 60)

sce
```

# Step 1: Filter sparse donors

Before aggregation, remove donor-cell type combinations with too few cells.

```{r filter}
sce <- filterSparseDonors(sce,
                           donor     = "donor",
                           cell_type = "cell_type",
                           min_cells = 5)
ncol(sce)
```

# Step 2: Inspect pseudo-bulk profiles

```{r pseudobulk}
pb <- fastPseudobulk(sce,
                      donor       = "donor",
                      cell_type   = "cell_type",
                      target_type = "CD4_Tcell")

cat("Pseudo-bulk matrix:", nrow(pb$pseudobulk),
    "genes x", ncol(pb$pseudobulk), "donors\n")

cat("\nDonor weights (sqrt of cell count):\n")
print(round(pb$donor_weights, 2))
```

Notice that all donors have equal cell counts here (10 each), so weights
are identical. In real data, weights will vary substantially.

# Step 3: Run differential expression

```{r fastde}
result <- fastDE(sce,
                  donor       = "donor",
                  cell_type   = "cell_type",
                  condition   = "condition",
                  target_type = "CD4_Tcell",
                  min_cells   = 5,
                  min_cpm     = 1,
                  min_donors  = 2)

result
```

```{r de_results}
# Top significant genes
dt <- as.data.frame(deTable(result))
dt_sorted <- dt[order(dt$adj.P.Val), ]
head(dt_sorted[, c("gene", "logFC", "P.Value", "adj.P.Val")], 15)
```

```{r de_summary}
# Summary
sig <- sum(dt$adj.P.Val < 0.05, na.rm = TRUE)
cat(sprintf("\n%d / %d genes significant (FDR < 0.05)\n",
            sig, nrow(dt)))
cat(sprintf("Injected genes recovered: %d / 20\n",
            sum(paste0("Gene", 1:20) %in%
                rownames(dt)[dt$adj.P.Val < 0.05])))
```

# Step 4: Visualise results

```{r volcano, fig.cap = "Volcano plot of DE results. Red = upregulated, blue = downregulated at FDR < 0.05 and |logFC| > 1."}
plotDEResults(result, fdr_thresh = 0.05, lfc_thresh = 1, top_n = 10)
```

# The donor weighting advantage

Here we demonstrate why weighting matters. We compare `fastDE` output
when one donor has far fewer cells than the others.

```{r weighting_demo}
# Create an unbalanced dataset
set.seed(99)
sce_unbal <- sce
# Simulate donor 1 having only 2 cells by keeping just 2
keep <- c(which(sce$donor != "Donor1"),
          which(sce$donor == "Donor1")[1:2])
sce_unbal <- sce_unbal[, keep]

pb_unbal <- fastPseudobulk(sce_unbal,
                             donor       = "donor",
                             cell_type   = "cell_type",
                             target_type = "CD4_Tcell")

cat("Donor weights (unbalanced):\n")
print(round(pb_unbal$donor_weights, 2))
```

Donor1's weight (`sqrt(2) ≈ 1.41`) is much lower than the other donors
(`sqrt(10) ≈ 3.16`), so its noisy pseudo-bulk has proportionally less
influence on the DE estimates — without being discarded entirely.

# Session information

```{r session}
sessionInfo()
```
